Creating optimal conditions for reproducible data analysis in \R with `fertile'\protect\thanks{The authors gratefully acknowledge contributions from Hadley Wickham, Jenny Bryan, Greg Wilson, Edgar Ruiz, and other members of the \pkg{tidyverse} team.}

### 2020-08-05 STAT-paper-template/APA/fertile.Rnw
revision

Data-based research is considered fully reproducible when the requisite code and data files
produce identical results when run by another analyst.

As research is becoming increasingly data-driven, and because knowledge can be shared worldwide so rapidly, reproducibility is critical to the advancement of scientific knowledge. Academics around the world have recognized this, and publications and discusions addressing reproducibility appear to have increased in the last several years (\cite{sep-scientific-reproducibility}, \cite{plos-biology}, \cite{bioessays-gosselin}, \cite{eisner-reproducibility}, \cite{engineering-reproducibility}).

@@

Existing attempts to address reproducibility in data science are often either too generalized, resulting in shallow and vague recommendations that are challenging to implement, or too specific, approaching one aspect of reproducibility well but doing so in a highly technical way failes to capture the bigger picture and creates challenges for inexperienced users.

@@

\section{Literature Review}\label{sec:lit}

\subsection{Written Work}\label{papers}

Papers and other published works focusing on reproducibility can be found in all areas of scientific research. However, as \cite{Goodman341ps12} argue, the language and conceptual framework of research reproducibility varies significantly across the sciences, and there are no clear standards on reproducibility agreed upon by the scientific community as a whole.

\pkg{fertile} considers recommendations from a variety of fields, attempting to determine the key aspects of reproducibility faced by scientists in different disciplines and address as many as possible in a single piece of software.

Although reproducibility recommendations vary, further investigation identifies several common ideas across publications.

@@

\begin{itemize}
\item[1] Use clear separation, labeling, and documentation.
\item[2] Automate processes when possible.
\item[3] Design the data analysis workflow as a sequence of small steps glued together, with
outputs from one step serving as inputs into the next. This is a common suggestion within the computing community, originating as part of the Unix philosophy \cite{unix}.
\end{itemize}

\cite{cooper2017guide} focuses on data analysis in \R and identifies a similar list of important reproducibility components and reiterates the need for clearly labeled, well-documented, and well-separated files. In addition, they recommend publishing a list of dependencies and using version control.

Broman reiterates the need for clear naming and file separation while sharing several additional suggestions: keep the project contained in one directory, use relative paths, and include a \cmd{README} (\cite{broman}).

@@

When considered in combination, these sources provide a well-rounded picture of the components important to research reproducibility. Using this literature as a guideline, we identify several key features of reproducible work to focus on. These recommendations are opinionated---due to the lack of agreement on which components of reproducibility are most important, we select those that are mentioned most often, as well as some that are mentioned less but that we view as important.

\begin{itemize}
\item[1] A well-designed file structure:
  \begin{itemize}
  \item Separate folders for different file types.
  \item No extraneous files.
  \item Minimal clutter.
  \end{itemize}
\item[2] Good documentation:
  \begin{itemize}
  \item Files are clearly named, preferably in a way where the order in which they should be run is clear.
  \item A README is present.
  \item Dependencies are noted.
  \end{itemize}
\item[3] Reproducible file paths:
  \begin{itemize}
  \item No absolute paths, or paths leading to locations outside of a project's directory, are used in code. Only portable (relative) paths.
  \end{itemize}
\item[4] Randomness is accounted for:
  \begin{itemize}
  \item If randomness is used in code, a seed must also be set.
  \end{itemize}
\item[5] Code conforms to \pkg{tidyverse} style:
  \begin{itemize}
  \item Although not discussed in any of the literature mentioned previously, we believe \pkg{tidyverse} integration to be very important. Code conformation to \pkg{tidyverse} style helps ensure that analyses meeting \pkg{fertile}'s recommendations will be able to integrate seamlessly with packages from the \pkg{tidyverse}, an ever-expanding collection of packages designed to operate together with one another.
  \end{itemize}
\end{itemize}

Much of the available literature focuses on file structure, organization, and naming, and \pkg{fertile}'s features are consistent with this. The ideal file structure is not agreed upon by academics, though there are some recommendations focused on \R that are available. \cite{marwick2018packaging} provides the framework for file structure that \pkg{fertile} is based on: a structure similar to that of an \R package (\cite{hadley-packages}, \cite{coreteam-extensions}), with an \cmd{R} folder, as well as \cmd{data}, \cmd{data-raw}, \cmd{inst}, and \cmd{vignettes}.

@@

Much of the written work discussed in Section \ref{papers} is highly generalized, written to be applicable to users working with a variety of statistical software programs. Because all statistical software programs operate differently, these recommendations are inherently vague and difficult to implement, particularly to new analysts who are relatively unfamiliar with their software.

Thus, attempts to address reproducibility that are specific to certain software programs are more likely to be successful. \R, due to its accessibility and popularity as a tool for statistical analysis, is an ideal candidate such recommendations.

Acknowledging this, researchers and members of the \R community have created a small body of \R packages that focus on research reproducibility, although not all of them are available on CRAN:

@@

Many of these packages are narrowly tailored, with each effectively addressing a small component of reproducibility: file structure, modularization of code, version control, etc. These packages often  succeed in their area of focus, but at the cost of accessibility to a wider audience. Their functions are often quite complex to use, and many steps must be completed to achieve the required reproducibility goal. This cumbersome nature means that most reproducibility packages currently available are not easily accessible to users near the beginning of their \R journey, nor particularly useful to those looking for quick and easy reproducibility checks.

A more effective way of realizing widespread reproducibility is to make the process for doing so simple enough that it takes little to no conscious effort to implement. You want users to "fall into a hole"\footnote{We paraphrase Hadley Wickham.} of good practice.

Several continuous integration code development tools external to \R attempt to do this, providing a way to address different aspects of reproducibility with minimal code. For example, \pkg{wercker}---a command line tool that leverages Docker---enables users to test whether their projects will successfully compile when run on a variety of operating systems without access to the user's local hard drive \cite{wercker}. GitHub Actions is integrated with GitHub and can be configured to do similar checks on projects hosted in repositories. \href{https://travis-ci.com}{Travis CI} and \href{https://circleci.com/product}{Circle CI} are popular continuous integration tools which can also be used to check \R code.

However, while these tools can be useful in some circumstances, they are generalized so as to be useful to the widest audience. As a result, their checks are not designed to be \R-specific, which makes them sub-optimal for users looking to address reproducibility issues involving features specific to the \R programming language, such as package installation and seed setting.

@@

\pkg{fertile} attempts to address these gaps in existing software by providing a simple, easy-to-learn reproducibility package that, rather than focusing intensely on a specific area, provides some information about a wide variety of aspects influencing reproducibility. \pkg{fertile} also allows for flexibility, offering benefits to users at any stage in the data analysis workflow, and provides R-specific features, which address certain aspects of reproducibility that can be missed by external project development software.

@@

<<fertile, message = FALSE, error = TRUE>>=
library(fertile)
search()
file.exists("~/Desktop/my_data.csv")
read.csv("~/Desktop/my_data.csv")
@

@@

Users can access the majority of \pkg{fertile}'s retroactive features through two primary functions, \func{proj\_check} and \func{proj\_analyze}.

The \func{proj\_check} function runs fifteen different reproducibility tests, noting which ones passed, which ones failed, the reason for failure, a recommended solution, and a guide to where to look for help. These tests include: looking for a clear build chain, checking to make sure the root level of the project is clear of clutter, confirming that there are no files present that are not being directly used by or created by the code, and looking for uses of randomness that to not have a call to \func{set.seed} present. A full list is provided below:

@@

@@

<<log-clear>>=
log_clear()
log_report()
@

\subsubsection{How It Works}

@@

\begin{enumerate}
\item Identify an \R function is likely to be involved in operations that may break reproducibility. Popular functions associated with only one package (e.g., \func{read\_csv} from \pkg{readr}) are ideal candidates.
\item Create a function in \pkg{fertile} with the same name, that takes the same arguments (and always the dots \cmd{...}).
\item Write this new function so that it: 1) captures any arguments; 2) logs the name of the function called; 3) perform any checks on these arguments; and finally 4) calls the original function with the original arguments. Except where warranted, the execution looks the same to the user as if they were calling the original function.
\end{enumerate}
Most shims are quite simple and look something like what is shown below for \func{read\_csv}.

@@

In order to ensure that the \pkg{fertile} versions of functions ("shims") are always favored over their original namesakes ("mask") when called, \pkg{fertile} uses its own shims of the \func{library} and \func{require} functions to  manipulate the \R \func{search} path so that it is always located in the first position. In the \pkg{fertile} version of \func{library}, we detach \pkg{fertile} from the search path, load the requested package, and then re-attach \pkg{fertile}. This ensures that when a user executes a command, \R will check \pkg{fertile} for a matching function before considering other packages.

@@

For users focused on file paths, the \func{check\_path} function analyzes a vector of paths (or a single path) to determine whether there are any absolute paths or paths that lead outside the project directory.

@@

<<sandbox>>=
dir <- getwd()
new_dir <- sandbox(dir)
dir
fs::dir_ls(dir) %>% head()
new_dir
fs::dir_ls(new_dir) %>% head()
@

@@

\pkg{fertile} is wide in scope. The package's simplicity enables users of any background to take advantage of its features, and its big-picture design gives \pkg{fertile} the potential to provide benefits across a variety of disciplines.

@@

She reads in the data, makes the graph, and knits her \cmd{.Rmd} file. It compiles successfully, so she submits the assignment. The next day, she receives an email from her professor saying that her assignment failed to compile and that she needs to make changes and try again.

Susan doesn't understand why it didn't work on the professor's computer when it did on her own. The professor recommends that she install \pkg{fertile} and run \func{proj\_check} on her assignment. She does this and gets a message informing her she used an absolute path to open her dataset, and she should use a relative path. She looks up what this means and then uses the new information to update her assignment. She resubmits and her second attempt is successful.

@@

Emma is a post-doc, with several years of \R experience. She is familiar with some basic rules of reproducibility--file paths should always be relative and randomness should always be associated with a seed--but has never needed to pass any sort of reproducibility check before because her professors never emphasized that.

@@

\pkg{fertile} is designed to:

\begin{enumerate}
\item Be simple enough that users with minimal \R experience can use the package without issue,
\item Increase the reproducibility of work produced by its users, and
\item Educate its users on why their work is or is not reproducible and provide guidance on how to address any problems.
\end{enumerate}

Anecdotal evidence and user feedback is useful, providing a baseline knowledge of whether these goals are being met. However, in order to truly test whether the package is successful in achieving its design goals, an experiment is necessary.

To test \pkg{fertile}'s effectiveness, we began an initial randomized control trial of the package on an introductory undergraduate data science course at Smith College in Spring 2020 \footnote{This study was approved by Smith College IRB, Protocol \#19-032}. Unfortunately, however, the trial had to be postponed after the COVID-19 pandemic significantly altered the experimental conditions at the midpoint of testing.

@@

\item The students are then asked to download a package called \pkg{sds192} (the course number and prefix), which was created for the purpose of this trial. It leverages an \func{.onAttach} function to scan the \R environment and collect the username of the user who is loading the package and run it through the same hashing algorithm as used previously. It then identifies whether that user belongs to the experimental or the control group. Depending on the group they are in, they receive a different version of the package:

@@

Although the experiment was unsuccessful in its first attempt, we hope to run the same trial again and gather data on \pkg{fertile}'s effectiveness.

@@

\pkg{fertile} is an \R package that lowers barriers to reproducible data analysis projects in \R, providing a wide array of checks and suggestions addressing many different aspects of project reproducibility, including file organization, file path usage, documentation, and dependencies.

\pkg{fertile} is meant to be educational, providing informative error messages that show how the ways in which the mistakes the user is making are problematic, as well as sharing recommendations on how to fix them. The package is designed in this way so as to promote a greater understanding of reproducibility concepts in its users, with the goal of increasing the overall awareness and understanding of reproducibility in the \R community.

@@

In the moment, it often feels easiest to take a shortcut--to use an absolute path or change a working directory. However, when considering the long term path of a project, spending the extra time to improve reproducibility is worthwhile. \pkg{fertile}'s user-friendly features can help data analysts avoid these harmful shortcuts with minimal effort.

@@

The process of ensuring true reproducibility is highly complex and challenging. There are a significant number of components, many of which are described in \ref{sec:lit}, which must all be in place for work to be reproducibile. \pkg{fertile}, in its current version, addresses the vast majority of these in some way. However, there are several areas where further development to extend the available features of the package would be beneficial. These include the following:

>>>>

Data-based research is considered fully reproducible when the requisite code and data files
produce identical results when run by another analyst. 
As research is becoming increasingly data-driven, and because knowledge can be shared worldwide so rapidly, reproducibility is critical to the advancement of scientific knowledge. Academics around the world have recognized this, and publications and discussions addressing reproducibility appear to have increased in the last several years (\cite{sep-scientific-reproducibility}, \cite{plos-biology}, \cite{bioessays-gosselin}, \cite{eisner-reproducibility}, \cite{engineering-reproducibility}).

@@

Existing attempts to address reproducibility in data science are often either too generalized---resulting in shallow and vague recommendations that are challenging to implement---or too specific---approaching one aspect of reproducibility well but doing so in a highly technical way that fails to capture the bigger picture and creates challenges for inexperienced users.

@@

\section{Related Work}\label{sec:lit}

\subsection{Literature Review}\label{sec:papers}

Publications on reproducibility can be found in all areas of scientific research. However, as \cite{Goodman341ps12} argue, the language and conceptual framework of research reproducibility varies significantly across the sciences, and there are no clear standards on reproducibility agreed upon by the scientific community as a whole.

We consider recommendations from a variety of fields, and determine the key aspects of reproducibility faced by scientists in different disciplines.

@@

\begin{enumerate}
\item Use clear separation, labeling, and documentation.
\item Automate processes when possible.
\item Design the data analysis workflow as a sequence of small steps glued together, with
outputs from one step serving as inputs into the next. This is a common suggestion within the computing community, originating as part of the Unix philosophy (\cite{unix}).
\end{enumerate}

\cite{cooper2017guide} focuses on data analysis in \R and identifies a similar list of important reproducibility components and reiterates the need for clearly labeled, well-documented, and well-separated files. In addition, they recommend publishing a list of dependencies and using version control.
\cite{broman} reiterates the need for clear naming and file separation while sharing several additional suggestions: keep the project contained in one directory, use relative paths, and include a \cmd{README}.

@@

When considered in combination, these sources provide a well-rounded picture of the components important to research reproducibility. Using this literature as a guideline, we identify several key features of reproducible work. These recommendations are opinionated---due to the lack of agreement on which components of reproducibility are most important, we select those that are mentioned most often, as well as some that are mentioned less but that we view as important.

\begin{enumerate}
\item A well-designed file structure:
  \begin{itemize}
  \item Separate folders for different file types.
  \item No extraneous files.
  \item Minimal clutter.
  \end{itemize}
\item Good documentation:
  \begin{itemize}
  \item Files are clearly named, preferably in a way where the order in which they should be run is clear.
  \item A README is present.
  \item Dependencies are noted.
  \end{itemize}
\item Reproducible file paths:
  \begin{itemize}
  \item No absolute paths, or paths leading to locations outside of a project's directory, are used in code---only portable (relative) paths.
  \end{itemize}
\item Randomness is accounted for:
  \begin{itemize}
  \item If randomness is used in code, a seed must also be set.
  \end{itemize}
\item Readable, styled code:
  \begin{itemize}
  \item Code should be written in a coherent style. Code that conforms to a style guide or is written in a consistent dialect is easier to read (\cite{hermans2017programming}). We believe that the \pkg{tidyverse} provides the most accessible dialect of \R.
  \end{itemize}
\end{enumerate}

Much of the available literature focuses on file structure, organization, and naming, and \pkg{fertile}'s features are consistent with this. 
\cite{marwick2018packaging} provide the framework for file structure that \pkg{fertile} is based on: a structure similar to that of an \R package (\cite{hadley-packages}, \cite{coreteam-extensions}), with an \cmd{R} folder, as well as \cmd{data}, \cmd{data-raw}, \cmd{inst}, and \cmd{vignettes}.

@@

Much of the work discussed in Section \ref{sec:papers} is highly generalized, written to be applicable to users working with a variety of statistical software programs. Because all statistical software programs operate differently, these recommendations are inherently vague and difficult to implement, particularly to new analysts who are relatively unfamiliar with their software.

Focused attempts to address reproducibility in specific certain software programs are more likely to be successful. 
We focus on \R, due to its open-source nature, accessibility, and popularity as a tool for statistical analysis.

A small body of \R packages focus on research reproducibility.

@@

Many of these packages are narrow, with each effectively addressing a small component of reproducibility: file structure, modularization of code, version control, etc. These packages often  succeed in their area of focus, but at the cost of accessibility to a wider audience. Their functions are often quite complex to use, and many steps must be completed to achieve the required reproducibility goal. This cumbersome nature means that most reproducibility packages currently available are not easily accessible to users near the beginning of their \R journey, nor particularly useful to those looking for quick and easy reproducibility checks.
A more effective way of realizing widespread reproducibility is to make the process for doing so simple enough that it takes little to no conscious effort to implement. You want users to "fall into a hole"\footnote{We paraphrase Hadley Wickham.} of good practice.

\emph{Continuous integration} tools provide more general approaches to automated checking; which can enhance reproducibility with minimal code. For example, \pkg{wercker}---a command line tool that leverages Docker---enables users to test whether their projects will successfully compile when run on a variety of operating systems without access to the user's local hard drive (\cite{wercker}). GitHub Actions is integrated into GitHub and can be configured to do similar checks on projects hosted in repositories. \href{https://travis-ci.com}{Travis CI} and \href{https://circleci.com/product}{Circle CI} are popular continuous integration tools which can also be used to check \R code.

However, while these tools can be useful, they are generalized so as to be useful to the widest audience. As a result, their checks are not designed to be \R-specific, which makes them sub-optimal for users looking to address reproducibility issues involving features specific to the \R programming language, such as package installation and seed setting.

\subsection{Our contribution}

\pkg{fertile} attempts to address these gaps in existing software by providing a simple, easy-to-learn reproducibility package that, rather than focusing intensely on a specific area, provides some information about a wide variety of aspects influencing reproducibility. \pkg{fertile} is flexible, offering benefits to users at any stage in the data analysis workflow, and provides \R-specific features, which address certain aspects of reproducibility that can be missed by external project development software.

@@

@@

<<fertile, message = FALSE, error = TRUE>>=
library(fertile)
file.exists("~/Desktop/my_data.csv")
read.csv("~/Desktop/my_data.csv")
read.csv("../../../Desktop/my_data.csv")
@

@@

Users can access the majority of \pkg{fertile}'s retroactive features through two primary functions, \func{proj\_check} and \func{proj\_analyze}.
The \func{proj\_check} function runs fifteen different reproducibility tests, noting which ones passed, which ones failed, the reason for failure, a recommended solution, and a guide to where to look for help. These tests include: looking for a clear build chain, checking to make sure the root level of the project is clear of clutter, confirming that there are no files present that are not being directly used by or created by the code, and looking for uses of randomness that do not have a call to \func{set.seed} present. A full list is provided below:

@@

\subsection{Logging}

@@

<<log-clear, include=FALSE>>=
log_clear()
log_report()
@

\subsection{How It Works}

@@

\begin{enumerate}
\item Identify an \R function that is likely to be involved in operations that may break reproducibility. Popular functions associated with only one package (e.g., \func{read\_csv} from \pkg{readr}) are ideal candidates.
\item Create a function in \pkg{fertile} with the same name that takes the same arguments (and always the dots \cmd{...}).
\item Write this new function so that it: 1) captures any arguments; 2) logs the name of the function called; 3) perform any checks on these arguments; and finally 4) calls the original function with the original arguments. Except where warranted, the execution looks the same to the user as if they were calling the original function.
\end{enumerate}
Most shims are quite simple and look something like what is shown below for \func{read\_csv}.

@@

In order to ensure that the \pkg{fertile} versions of functions ("shims") are always favored ("mask") over their original namesakes when called, \pkg{fertile} uses its own shims of the \func{library} and \func{require} functions to  manipulate the \R \func{search} path so that it is always located in the first position. In the \pkg{fertile} version of \func{library}, we detach \pkg{fertile} from the search path, load the requested package, and then re-attach \pkg{fertile}. This ensures that when a user executes a command, \R will check \pkg{fertile} for a matching function before considering other packages.

@@

The \func{check\_path} function analyzes a vector of paths (or a single path) to determine whether there are any absolute paths or paths that lead outside the project directory.

@@

<<sandbox>>=
dir <- getwd()
new_dir <- sandbox(dir)
dir
fs::dir_ls(dir) %>% head(3)
new_dir
fs::dir_ls(new_dir) %>% head(3)
@

@@

\pkg{fertile}'s simplicity enables users of any background to take advantage of its features, and its big-picture design gives \pkg{fertile} the potential to provide benefits across a variety of disciplines.

@@

She reads in the data, makes the graph, and knits her \cmd{.Rmd} file. It compiles successfully, so she submits the assignment. The next day, she receives an email from her professor saying that her assignment failed to compile and that she needs to make changes and try again.
Susan doesn't understand why it didn't work on the professor's computer when it did on her own. The professor recommends that she install \pkg{fertile} and run \func{proj\_check} on her assignment. She does this and gets a message informing her she used an absolute path to open her dataset, and she should use a relative path. She looks up what this means and then uses the new information to update her assignment. She resubmits and her second attempt is successful.

@@

Emma is a post-doc, with several years of \R experience. She is familiar with some basic rules of reproducibility---file paths should always be relative and randomness should always be associated with a seed---but has never needed to pass any sort of reproducibility check before because her professors never emphasized that.

@@

\pkg{fertile} is designed to: 1) be simple enough that users with minimal \R experience can use the package without issue; 2) increase the reproducibility of work produced by its users, and; 3) educate its users on why their work is or is not reproducible and provide guidance on how to address any problems.

To test \pkg{fertile}'s effectiveness, we began an initial randomized control trial of the package on an introductory undergraduate data science course at Smith College in Spring 2020 \footnote{This study was approved by Smith College IRB, Protocol \#19-032}. Unfortunately, the trial had to be postponed after the COVID-19 pandemic significantly altered the experimental conditions at the midpoint of testing.
Although the experiment was unsuccessful in its first attempt, we hope to run the same trial again and gather data on \pkg{fertile}'s effectiveness.

@@

\item The students are then asked to download a package called \pkg{sds192} (the course number and prefix), which was created for the purpose of this trial. It leverages an \func{.onAttach} function to scan the \R environment and collect the username of the user who is loading the package and run it through the same hashing algorithm as used previously. It then identifies whether that user belongs to the experimental or the control group. Depending on the group they are in, they receive a different version of the package.

@@

@@

\pkg{fertile} is an \R package that lowers barriers to reproducible data analysis projects in \R, providing a wide array of checks and suggestions addressing many different aspects of project reproducibility, including file organization, file path usage, documentation, and dependencies.
\pkg{fertile} is meant to be educational, providing informative error messages that show how the ways in which the mistakes the user is making are problematic, as well as sharing recommendations on how to fix them. The package is designed in this way so as to promote a greater understanding of reproducibility concepts in its users, with the goal of increasing the overall awareness and understanding of reproducibility in the \R community.

@@

In the moment, it often feels easiest to take a shortcut---to use an absolute path or change a working directory. However, when considering the long term path of a project, spending the extra time to improve reproducibility is worthwhile. \pkg{fertile}'s user-friendly features can help data analysts avoid these harmful shortcuts with minimal effort.

@@

\pkg{fertile}, in its current version, addresses the vast majority of the aspects of reproducibility identified in Section \ref{sec:papers} in some way. However, there are several areas where further development to extend the available features of the package would be beneficial. These include the following:
